Fractional Types

### 2016-12-15 typespp.tex
clean up more and up to sec 2 small revisions

\usepackage{amsthm,amstext,amssymb,amsmath}
\usepackage{datetime}
\usepackage{bbold}
\usepackage{url}
\usepackage{xcolor}
\usepackage{multicol}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{bussproofs}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{quotes}
\usepackage[section]{placeins}
\usepackage{fdsymbol}

\newcommand{\amr}[1]{\fbox{\begin{minipage}{0.4\textwidth}\color{purple}{Amr says: #1}\end{minipage}}}
\newcommand{\vic}[1]{\fbox{\begin{minipage}{0.4\textwidth}\color{purple}{Vikraman says: #1}\end{minipage}}}
\newcommand{\jc}[1]{\fbox{\begin{minipage}{\textwidth}\color{purple}{Jacques says: #1}\end{minipage}}}

\newcommand{\hide}[1]{}

\newcommand{\textpi}{$\pi$}
\newcommand{\ft}{\mathbb{T}}
\newcommand{\hash}{\#}
\newcommand{\dg}{conjugation groupoid}
\newcommand{\DG}{Conjugation Groupoid}
\newcommand{\pifrac}{\ensuremath{\Pi^/}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\isotwo}{\Leftrightarrow}
\newcommand{\alt}{~|~}
\newcommand{\ag}[2]{#1 \sslash #2}
\newcommand{\agp}[3]{#1 \sslash_{#3} #2}
\newcommand{\order}[1]{\hash #1}
\newcommand{\iorder}[1]{1/\hash #1}
\newcommand{\oneg}[1]{\mathbb{1}_{#1}}
\newcommand{\divgl}[2]{\hash{#1} \div_{l} \hash{#2}}
\newcommand{\divgr}[2]{\hash{#1} \div_{r} \hash{#2}}
\newcommand{\ord}[1]{\ensuremath{\mathsf{order}(#1)}}
\newcommand{\inl}[1]{\textsf{inl}(#1)}
\newcommand{\inr}[1]{\textsf{inr}(#1)}
\newcommand{\zt}{\mathbb{0}}
\newcommand{\ot}{\mathbb{1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Zn}{\mathbb{Z}_n}
\newcommand{\Zvn}{\mathbb{Z}^v_n}
\newcommand{\N}{\mathbb{N}}
\newcommand{\cycle}{\textsf{cycle}}
\newcommand{\twod}{\mathbb{T}}
\newcommand{\fract}[2]{#1/#2}
\newcommand{\fv}[2]{\fcolorbox{black}{white}{\strut $#1$}\fcolorbox{black}{gray!20}{$\strut #2$}}
\newcommand{\pt}[2]{\bullet[#1,#2]}
\newcommand{\sing}[1]{\textsc{Sing}(#1)}
\newcommand{\singi}[2]{\textsc{SingI}_{{#1}}(#2)}
\newcommand{\iter}[1]{\textsc{Iter}(#1)}
\newcommand{\pair}[2]{\langle #1,#2 \rangle}
\newcommand{\triple}[3]{\langle #1,#2,#3 \rangle}
\newcommand{\distiterplus}[3]{\mathsf{dist}~#1~#2~#3}

@@

\keywords{dummy}
\subjclass{dummy}

@@

\begin{abstract}

We exhibit types whose natural cardinality is fractional.  More
precisely, we show that the groupoid cardinality (as defined by
Baez-Dolan) of the denotation of the type of a singleton reversible
program $p$ with exactly $k$ distinct proofs of reversibility has
cardinality $1/k$.  We further show that this type is naturally a
multiplicative inverse to the type of all iterates $p ^ i$ of that
reversible program.

We situate this work as an extension of a larger reversible
programming language ($\Pi$), and show that this extension is also
reversible.  Interestingly, this extension supports first-class
functions as well as some natural analogues of operations coming from
compact closed categories.  We emphasize that the key ingredients are
reversibility and proof relevance: cardinality $1/k$ arises from
having exactly $k$ proofs of reversibility.

\end{abstract}

@@

\begin{center}
\begin{tabular}{c@{\kern -4pt}c}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]
  \draw[dashed] (0,0) ellipse (3cm and 1.2cm);
  \node[below] (A) at (-1.7,0.3) {\texttt{a}};
  \node[below] (B) at (-1.1,0.3) {\texttt{b}};
  \node[below] (C) at (-0.5,0.3) {\texttt{c}};
  \node[below] (D) at (0.5,0.3) {\texttt{d}};
  \node[below] (E) at (1.1,0.3) {\texttt{e}};
  \node[below] (F) at (1.7,0.3) {\texttt{f}};
  \draw[fill] (-1.7,0.3) circle [radius=0.05];
  \draw[fill] (-1.1,0.3) circle [radius=0.05];
  \draw[fill] (-0.5,0.3) circle [radius=0.05];
  \draw[fill] (0.5,0.3) circle [radius=0.05];
  \draw[fill] (1.1,0.3) circle [radius=0.05];
  \draw[fill] (1.7,0.3) circle [radius=0.05];
  \draw (-1.7,0.3) -- (-1.1,0.3);
  \draw (-1.1,0.3) -- (-0.5,0.3);
  \draw (-1.7,0.3) to[bend left] (-0.5,0.3) ;
  \draw (0.5,0.3) -- (1.1,0.3);
  \draw (1.1,0.3) -- (1.7,0.3);
  \draw (0.5,0.3) to[bend left] (1.7,0.3) ;
  \path (A) edge [loop below] node[below] {\texttt{id}} (B);
  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
  \path (C) edge [loop below] node[below] {\texttt{id}} (B);
  \path (D) edge [loop below] node[below] {\texttt{id}} (B);
  \path (E) edge [loop below] node[below] {\texttt{id}} (B);
  \path (F) edge [loop below] node[below] {\texttt{id}} (B);
\end{tikzpicture}
&
\begin{tikzpicture}[scale=0.5,every node/.style={scale=0.5}]
  \draw[dashed] (0,0) ellipse (2cm and 2.4cm);
  \node[below] (B) at (0,-1.1) {\texttt{*}};
  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
  \path (B) edge [loop above, looseness=15, in=48, out=132] node[above] {$p$} (B);
  \path (B) edge [loop above, looseness=25, in=40, out=140] node[above] {$q$} (B);
  \path (B) edge [loop above, looseness=40, in=32, out=148] node[above] {$q'$} (B);
  \path (-1,0.4) edge[double, bend left] node[left] {$\alpha$} (-1,1.0);
\end{tikzpicture}
\end{tabular}
\end{center}

\medskip\noindent Baez and Dolan~\cite{groupoidcard} assign to each groupoid a
\emph{cardinality} that counts the objects up to equivalences. The
groupoid on the left has six points \texttt{a}, \texttt{b}, \texttt{c},
\texttt{d}, \texttt{e}, and \texttt{f} with two groups of three
points clustered in an equivalence class and hence the groupoid has
cardinality~2. The (2-)groupoid on the right has one point $\ast$ with
four equivalences \texttt{id}, $p$, $q$, and~$q'$ on it. The
equivalences $q$ and $q'$ are however identified by~$\alpha$ leaving
only three distinct equivalence classes and hence making the
cardinality $\frac{1}{3}$.

@@

For our example combinators on the type $\mathbb{3}$, simple traces using the
operational semantics show the combinator $\permone$ is the identity
permutation; the combinators $\permthree$ and $\permsix$ swap two of the three
elements leaving the third intact; and the combinators $\permfour$ and
$\permfive$ rotate the three elements. We therefore have:
\[\begin{array}{rcl}
\mathit{order}(\permone) &=& 1 \\
\mathit{order}(\permtwo) = \mathit{order}(\permthree) = \mathit{order}(\permsix) &=& 2 \\
\mathit{order}(\permfour) = \mathit{order}(\permfive) &=& 3
\end{array}\]
We should note that the above definition is the only one in this
paper which is not \emph{effective};  in other words, we do not have an
algorithm for computing it.  While there is an obvious method to
compute it using the action of a 1-combinator on the elements of the
type it acts on, this is extremely inefficient.  What we mean is that
we know of no effective algorithm which works on the syntax of
combinators.  The (only) difficulty is $\odot$, which can have an
arbitrary effect on the order.

@@

\begin{figure}[t]
\begin{center}
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (1.5cm and 2.1cm);
  \node[below] (A) at (-0.5,0) {$a$};
  \node[below] (C) at (0.5,0) {$c$};
  \path (A) edge [loop below] node[below] {\texttt{id}} (A);
  \path (C) edge [loop below] node[below] {\texttt{id}} (C);
  \path (C) edge [out=140, in=40, looseness=4] (C);
\end{tikzpicture}
\qquad \qquad \qquad
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (3cm and 2.5cm);
  \node[below] (A) at (-2,0) {$a$};
  \node[below] (B) at (0,0) {$b$};
  \node[below] (C) at (1.5,0) {$c$};
  \path (A) edge [bend left=50] (B);
  \path (C) edge [out=140, in=40, looseness=4] (C);
  \path (A) edge [loop below] node[below] {\texttt{id}} (A);
  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
  \path (C) edge [loop below] node[below] {\texttt{id}} (C);
\end{tikzpicture}
\qquad \qquad \qquad
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (2.8cm and 2.5cm);
  \node[below] (A) at (-1.6,0) {$a$};
  \node[below] (B) at (0,0) {$b$};
  \node[below] (C) at (1.6,0) {$c$};
  \path (A) edge [loop below] node[below] {\texttt{id}} (A);
  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
  \path (C) edge [loop below] node[below] {\texttt{id}} (C);
  \path (A) edge [loop above, looseness=3, in=48, out=132] (A);
  \path (B) edge [loop above, looseness=3, in=48, out=132] (B);
  \path (C) edge [loop above, looseness=3, in=48, out=132] (C);
\end{tikzpicture}
\end{center}
\caption{\label{fig:groupoids2}Example groupoids $G_1$, $G_2$, and $G_3$.}
\end{figure}

>>>>

\usepackage{amsthm,amstext,amssymb,amsmath}
\usepackage{datetime}
\usepackage{fdsymbol} %% must be loaded before bbold because it interferes with numbers
\usepackage{bbold}
\usepackage{url}
\usepackage{xcolor}
\usepackage{multicol}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{bussproofs}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{quotes}
\usepackage[section]{placeins}

\newcommand{\textpi}{$\pi$}
\newcommand{\ft}{\mathbb{T}}
\newcommand{\hash}{\#}
\newcommand{\pifrac}{\ensuremath{\Pi^/}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\isotwo}{\Leftrightarrow}
\newcommand{\alt}{~|~}
\newcommand{\ag}[2]{#1 \sslash #2}
\newcommand{\agp}[3]{#1 \sslash_{#3} #2}
\newcommand{\order}[1]{\hash #1}
\newcommand{\iorder}[1]{1/\hash #1}
\newcommand{\oneg}[1]{\mathbb{1}_{#1}}
\newcommand{\divgl}[2]{\hash{#1} \div_{l} \hash{#2}}
\newcommand{\divgr}[2]{\hash{#1} \div_{r} \hash{#2}}
\newcommand{\ord}[1]{\ensuremath{\mathsf{order}(#1)}}
\newcommand{\inl}[1]{\textsf{inl}(#1)}
\newcommand{\inr}[1]{\textsf{inr}(#1)}
\newcommand{\zt}{\mathbb{0}}
\newcommand{\ot}{\mathbb{1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Zn}{\mathbb{Z}_n}
\newcommand{\Zvn}{\mathbb{Z}^v_n}
\newcommand{\N}{\mathbb{N}}
\newcommand{\cycle}{\textsf{cycle}}
\newcommand{\twod}{\mathbb{T}}
\newcommand{\fract}[2]{#1/#2}
\newcommand{\fv}[2]{\fcolorbox{black}{white}{\strut $#1$}\fcolorbox{black}{gray!20}{$\strut #2$}}
\newcommand{\pt}[2]{\bullet[#1,#2]}
\newcommand{\sing}[1]{\textsc{Sing}(#1)}
\newcommand{\singi}[2]{\textsc{SingI}_{{#1}}(#2)}
\newcommand{\iter}[1]{\textsc{Iter}(#1)}
\newcommand{\pair}[2]{\langle #1,#2 \rangle}
\newcommand{\triple}[3]{\langle #1,#2,#3 \rangle}
\newcommand{\distiterplus}[3]{\mathsf{dist}~#1~#2~#3}

@@

@@

\begin{abstract}
We exhibit types whose natural cardinality is fractional.  More
precisely, we show that the groupoid cardinality (as defined by
Baez-Dolan) of the denotation of the type of a singleton reversible
program $p$ with exactly $k$ distinct proofs of reversibility has
cardinality $1/k$.  We further show that this type is naturally a
multiplicative inverse to the type of all iterates $p ^ i$ of that
reversible program.

We situate this work as an extension of a larger reversible
programming language ($\Pi$), and show that this extension is also
reversible.  Interestingly, this extension supports first-class
functions as well as some natural analogues of operations coming from
compact closed categories.  We emphasize that the key ingredients are
reversibility and proof relevance: cardinality $1/k$ arises from
having exactly $k$ proofs of reversibility.
\end{abstract}

@@

\begin{center}
\begin{tabular}{c@{\qquad\qquad}c}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]
  \draw[dashed] (0,0) ellipse (3cm and 1.5cm);
  \node[below] (A) at (-1.7,0.3) {\texttt{a}};
  \node[below] (B) at (-1.1,0.3) {\texttt{b}};
  \node[below] (C) at (-0.5,0.3) {\texttt{c}};
  \node[below] (D) at (0.5,0.3) {\texttt{d}};
  \node[below] (E) at (1.1,0.3) {\texttt{e}};
  \node[below] (F) at (1.7,0.3) {\texttt{f}};
  \draw[fill] (-1.7,0.3) circle [radius=0.05];
  \draw[fill] (-1.1,0.3) circle [radius=0.05];
  \draw[fill] (-0.5,0.3) circle [radius=0.05];
  \draw[fill] (0.5,0.3) circle [radius=0.05];
  \draw[fill] (1.1,0.3) circle [radius=0.05];
  \draw[fill] (1.7,0.3) circle [radius=0.05];
  \draw (-1.7,0.3) -- (-1.1,0.3);
  \draw (-1.1,0.3) -- (-0.5,0.3);
  \draw (-1.7,0.3) to[bend left] (-0.5,0.3) ;
  \draw (0.5,0.3) -- (1.1,0.3);
  \draw (1.1,0.3) -- (1.7,0.3);
  \draw (0.5,0.3) to[bend left] (1.7,0.3) ;
  \path (A) edge [my loop] node[below] {\texttt{id}} (B);
  \path (B) edge [my loop] node[below] {\texttt{id}} (B);
  \path (C) edge [my loop] node[below] {\texttt{id}} (B);
  \path (D) edge [my loop] node[below] {\texttt{id}} (B);
  \path (E) edge [my loop] node[below] {\texttt{id}} (B);
  \path (F) edge [my loop] node[below] {\texttt{id}} (B);
\end{tikzpicture}
&
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.6}]
  \draw[dashed] (0,0) ellipse (2cm and 2.6cm);
  \node[below] (B) at (0,-1.1) {\texttt{*}};
  \path (B) edge [my loop] node[below] {\texttt{id}} (B);
  \path (B) edge [loop above, looseness=15, in=48, out=132] node[above] {$p$} (B);
  \path (B) edge [loop above, looseness=25, in=40, out=140] node[above] {$q$} (B);
  \path (B) edge [loop above, looseness=40, in=32, out=148] node[above] {$q'$} (B);
  \path (-1,0.4) edge[double, bend left] node[left] {$\alpha$} (-1,1.0);
\end{tikzpicture}
\end{tabular}
\end{center}

\medskip\noindent Baez and Dolan~\cite{groupoidcard} assign to each groupoid a
\emph{cardinality} that counts the objects up to equivalences. The
groupoid on the left has six points \texttt{a}, \texttt{b}, \texttt{c},
\texttt{d}, \texttt{e}, and \texttt{f} with two groups of three
points each clustered in an equivalence class and hence the groupoid has
cardinality~2. The (2-)groupoid on the right has one point $\ast$ with
four equivalences \texttt{id}, $p$, $q$, and~$q'$ on it. The
equivalences $q$ and $q'$ are however identified by~$\alpha$ leaving
only three distinct equivalence classes and hence making the
cardinality $\frac{1}{3}$.

@@

For our example combinators on the type $\mathbb{3}$, simple traces using the
operational semantics show the combinator $\permone$ is the identity
permutation; the combinators $\permthree$ and $\permsix$ swap two of the three
elements leaving the third intact; and the combinators $\permfour$ and
$\permfive$ rotate the three elements. We therefore have:
\[\begin{array}{rcl}
\mathit{order}(\permone) &=& 1 \\
\mathit{order}(\permtwo) = \mathit{order}(\permthree) = \mathit{order}(\permsix) &=& 2 \\
\mathit{order}(\permfour) = \mathit{order}(\permfive) &=& 3
\end{array}\]
We should note that the above definition is the only one in this paper which is not \emph{effective}. While there is an obvious method to compute it using the action of a 1-combinator on the elements of the type it acts on, this is extremely inefficient. We do not have an effective algorithm for computing it that works on the syntax of combinators.  The (only) difficulty is $\odot$, which can have an arbitrary effect on the order.

@@

\begin{figure}[t]
\begin{center}
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (1.5cm and 2.1cm);
  \node[below] (A) at (-0.5,0) {$a$};
  \node[below] (C) at (0.5,0) {$c$};
  \path (A) edge [my loop] node[below] {\texttt{id}} (A);
  \path (C) edge [my loop] node[below] {\texttt{id}} (C);
  \path (C) edge [out=140, in=40, looseness=4] (C);
\end{tikzpicture}
\qquad \qquad \qquad
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (3cm and 2.5cm);
  \node[below] (A) at (-2,0) {$a$};
  \node[below] (B) at (0,0) {$b$};
  \node[below] (C) at (1.5,0) {$c$};
  \path (A) edge [bend left=50] (B);
  \path (C) edge [out=140, in=40, looseness=4] (C);
  \path (A) edge [my loop] node[below] {\texttt{id}} (A);
  \path (B) edge [my loop] node[below] {\texttt{id}} (B);
  \path (C) edge [my loop] node[below] {\texttt{id}} (C);
\end{tikzpicture}
\qquad \qquad \qquad
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (2.8cm and 2.5cm);
  \node[below] (A) at (-1.6,0) {$a$};
  \node[below] (B) at (0,0) {$b$};
  \node[below] (C) at (1.6,0) {$c$};
  \path (A) edge [my loop] node[below] {\texttt{id}} (A);
  \path (B) edge [my loop] node[below] {\texttt{id}} (B);
  \path (C) edge [my loop] node[below] {\texttt{id}} (C);
  \path (A) edge [loop above, looseness=3, in=48, out=132] (A);
  \path (B) edge [loop above, looseness=3, in=48, out=132] (B);
  \path (C) edge [loop above, looseness=3, in=48, out=132] (C);
\end{tikzpicture}
\end{center}
\caption{\label{fig:groupoids2}Example groupoids $G_1$, $G_2$, and $G_3$.}
\end{figure}
